Topological Equivariant Artist Model for Visualization Library Architecture

### 2024-12-03 paper/main.tex
mostly acknowledgments & rework of 2

\title{Topological Equivariant Artist Model for Visualization Library Architecture}
% author names and IEEE memberships
\author{Hannah~Aizenman, Mikael Vejdemo-Johansson, Thomas~Caswell, and~Michael~Grossberg,~\IEEEmembership{Member,~IEEE,}% <-this % stops a space
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem H. Aizenman and M. Grossberg are with the department of Computer Science, City College of New York.
\protect\\
% note need leading \protect in front of \\ to get a newline within \thanks as
% \\ is fragile and will error, could use \hfil\break instead.
E-mail: haizenman@.edu, mgrossberg@ccny.cuny.edu
\IEEEcompsocthanksitem Mikael Vejdemo-Johansson is with the department of Mathematics, CUNY College of Staten Island.
\protect\\
E-mail: mvj@math.csi.cuny.edu
\IEEEcompsocthanksitem Thomas Caswell is with National Synchrotron Light Source II, Brookhaven National Lab
\protect \\
E-mail: tcaswell@bnl.gov}% <-this % stops an unwanted space
\thanks{Manuscript received X XX, XXXX; revised X XX, XXXX.}
}

@@

\IEEEPARstart{V}isualization design guidelines, generally, describe how to choose visual encodings that preserve the structure of the data; to follow these guidelines the visualization tools that implement these $\data \rightarrow graphic$ transforms must be structure preserving. Loosely, preserving structure means that the athematical properties of the data and how the points are connected to each other should be inferable from the graphic such that  a $graphic \rightarrow data$ mapping can be made. For example, values read off a bar chart have to be equivalent to the values used to construct that chart. Therefore a visualization tool is structure preserving when it preserves the bidirectional mapping $data\leftrightarrow graphic$.

@@

Since the earliest maps \cite{friendlyBriefHistoryData2006}, a visualization has been expected to preserve the $field$ properties and $topology$ of the corresponding dataset:
\begin{LaTeXdescription}
  \item[\textcolor{fiber}{\textbf{field}} \footnote{Throughout this paper, we augment the math with color to group conceptually related terms\cite{headMathAugmentationHow2022}, for example \textcolor{fiber}{field} and \textcolor{fiber}{fiber}. This color coding carries through to the figures.}]  is a set of values of the same type, e.g. one column of a table or the pixels of an image
  \item[\textcolor{base}{\textbf{topology}}] is the connectivity and relative positioning of elements in a dataset \cite{wilkinsonGrammarGraphics2005}.
\end{LaTeXdescription}

@@

While Bertin and Mackinlay's work primarily discusses the $data \rightarrow graphic$, Kindlemann and Scheidegger's algebraic framework \cite{kindlmannAlgebraicProcessVisualization2014} also dictates a set of conditions under which the $graphic \rightarrow data$  mapping is structure preserving. Encapsulating their conditions, we: present a uniform abstract data representation layer in \autoref{sec:atct:sheaves} for ensuring that the visualization should not change if the data representation (i.e. the data container) changes, define the conditions under which data is mapped unambiguously to visual encodings \cite{ziemkiewiczEmbeddingInformationVisualization2009} in \autoref{sec:atct:xi}, and provide a methodology for verifying that changes in data should correspond to changes in the visualization in \autoref{sec:artist:equivariant:artist} that does not necessarily require that the changes be perceptually significant. Furthermore, our model extends their conditions of equivariance by explicitly incorporating topology and by providing a framework for translating the theoretical ideas into buildable components in \autoref{sec:construction}.

@@

@@

\subsection{Structure Preservation In Software}
\label{sec:related-work:software}
Visualization libraries are in part measured by how expressive the components of the library are, where expressiveness is a measure of which structure preserving mappings a tool can implement \cite{mackinlayAutomatingDesignGraphical1986}. While some visualization tools aim to automate the pairing of data with structure preserving visual representations, such as Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007}, many visualization libraries leave that choice to the user. For example, connectivity assumptions tend to be embedded in each of the visual algorithms of `building block` libraries, a term used by Wongsuphasawat \cite{wongsuphasawatNavigatingWideWorld2021} to describe libraries that provide modular components for building elements of a visualization, such as functions for making boxes or translating data values to colors. In building block libraries such as Matplotlib\cite{hunterMatplotlib2DGraphics2007} and D3\cite{bostockDataDrivenDocuments2011} the connectivity assumptions lead to very different interfaces; for example in Matplotlib methods for updating data and parameters for controlling aesthetics differ between plots. While VTK\cite{hanwellVisualizationToolkitVTK2015,geveciVTK2012} provides a language for expressing the topological properties of the data, and therefore can embed that information in its visual algorithms, it does so in a non-uniform manner. On the other hand, domain specific libraries are designed with the assumption of continuities that are common in the domain \cite{HeerSoftware2006}, and therefore can somewhat restrict the interface to choices that that are appropriate for the domain. For example, a tabular topological structure of discrete rows, as illustrated in \autoref{fig:related-work:continuity:ktypes}, is assumed by
A Presentation Tool\cite{mackinlayAutomatingDesignGraphical1986, mackinlayAutomatingDesignGraphical1986} and grammar of graphics\cite{wilkinsonGrammarGraphics2005} and the ggplot\cite{wickhamGgplot2ElegantGraphics2016}, vega\cite{satyanarayanDeclarativeInteractionDesign2014}, and altair\cite{vanderplasAltairInteractiveStatistical2018} libraries built on these frameworks. Image libraries such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and its humanities ImagePlot\cite{studiesCulturevisImageplot2021} plugin assume that the input is 2D continuous. Networking libraries such as gephi\cite{bastianGephiOpenSource2009} and networkx\cite{HagbergExploringNetwork2008} assume a graph-like structure. By assuming the structure of their data, these domain specific libraries can provide more cohesive interfaces for a much more limited set of visualization algorithms than the building block libraries offer.
\note{transition sentence}
In this work, we formally specify the conditions necessary to build structure preserving components using category theory, because it provides a rich language for expressing the structure of objects, functions between objects, and the constraints that must hold for these functions to compose \cite{wielsManagementEvolvingSpecifications1998,yorgeyMonoidsThemeVariations2012}. A brief visualization oriented introduction to category theory is in Vickers et al \cite{vickersUnderstandingVisualizationFormal2013}, but they are applying category theory to semantic concerns about visualization design rather than library architecture.

@@

\section*{Acknowledgment}
\note{Acknowledge all the actual people}
The authors would like to thank...

Hannah is also very grateful to the nlab and wikipedia contributors who wrote such
clear explainations of many of the topics discussed in this paper.

>>>>

\title{Topological Equivariant Artist Model for Visualization Library Architecture}
% author names and IEEE memberships
\author{Hannah~Aizenman, Mikael Vejdemo-Johansson, Thomas~Caswell, and~Michael~Grossberg,~\IEEEmembership{Member,~IEEE,}% <-this % stops a space
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem H. Aizenman is with the department of Computer Science, The Gradaute Center, CUNY.
\protect\\
% note need leading \protect in front of \\ to get a newline within \thanks as
% \\ is fragile and will error, could use \hfil\break instead.
E-mail: haizenman@gradcenter.cuny.edu,
\IEEEcompsocthanksitem M. Grossberg is with the department of Computer Science, City College of New York, CUNY.
E-mail: mgrossberg@ccny.cuny.edu
\IEEEcompsocthanksitem Mikael Vejdemo-Johansson is with the department of Mathematics, CUNY College of Staten Island.
\protect\\
E-mail: mvj@math.csi.cuny.edu
\IEEEcompsocthanksitem Thomas Caswell is with National Synchrotron Light Source II, Brookhaven National Lab
\protect \\
E-mail: tcaswell@bnl.gov}% <-this % stops an unwanted space
\thanks{Manuscript received X XX, XXXX; revised X XX, XXXX.}
}

@@

\IEEEPARstart{V}isualization design guidelines, generally, describe how to choose visual encodings that preserve the structure of the data; to follow these guidelines the visualization tools that implement these $\data \rightarrow graphic$ transforms must be structure preserving. Loosely, preserving structure means that the properties of the data and how the points are connected to each other should be inferable from the graphic such that a $graphic \rightarrow data$ mapping can be made. For example, values read off a bar chart have to be equivalent to the values used to construct that chart. Therefore a visualization tool is structure preserving when it preserves the bidirectional mapping $data\leftrightarrow graphic$.

@@

In this section, we describe how structure has traditionally been discussed in visualization and mathematics and encapsulated in visualization library design because we propose that by formally incorporating mathematical definitions of equality, we provide a general framework that can account for a larger variety of fields than the existing work has, can formally expresse what it means to preserve the topology of the data, and we can do so in a uniform manner using category theory because that can lead to more consistent software interfaces in visualization software libraries \cite{wielsManagementEvolvingSpecifications1998,yorgeyMonoidsThemeVariations2012}.

Generally, preserving structure means that a visualization is expected to preserve the $field$ properties and $topology$ of the corresponding dataset:

\begin{LaTeXdescription}
  \item[\textcolor{fiber}{\textbf{field}} \footnote{Throughout this paper, we use color in definitions, equations, and visualizations to group conceptually related terms\cite{headMathAugmentationHow2022}, for example \textcolor{fiber}{field} and \textcolor{fiber}{fiber}.}] is a set of values of the same type, e.g. one column of a table or the pixels of an image
  \item[\textcolor{base}{\textbf{topology}}] is the connectivity and relative positioning of elements in a dataset \cite{wilkinsonGrammarGraphics2005}.
\end{LaTeXdescription}

The conditions under which $data \rightarrow graphic$ is structure preserving is discussed extensively in the visualization literature, codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011} and extended to tool design by Mackinlay\cite{mackinlayAutomaticDesignGraphical1987}, and a set of conditions under which the $graphic \rightarrow data$ mapping is structure preserving is presented in Kindlemann and Scheidegger's algebraic framework \cite{kindlmannAlgebraicProcessVisualization2014}. Encapsulating their conditions, we present a uniform abstract data representation layer in \autoref{sec:atct:sheaves} for ensuring that the visualization should not change if the data representation (i.e. the data container) changes, define the conditions under which data is mapped unambiguously to visual encodings \cite{ziemkiewiczEmbeddingInformationVisualization2009} in \autoref{sec:atct:xi}, and provide a methodology for verifying that changes in data should correspond to changes in the visualization in \autoref{sec:artist:equivariant:artist} that does not necessarily require that the changes be perceptually significant. Furthermore, our model extends their conditions of equivariance by explicitly incorporating topology and by providing a framework for translating the theoretical ideas into buildable components in \autoref{sec:construction}.

@@

@@

A brief visualization oriented introduction to category theory is in Vickers et al \cite{vickersUnderstandingVisualizationFormal2013}, but they are applying category theory to semantic concerns about visualization design rather than library architecture.

@@

\subsection{Structure Preservation In Software}
\label{sec:related-work:software}
Visualization libraries are in part measured by how expressive the components of the library are, where expressiveness is a measure of which structure preserving mappings a tool can implement \cite{mackinlayAutomatingDesignGraphical1986}. While some visualization tools aim to automate the pairing of data with structure preserving visual representations, such as Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007}, many visualization libraries leave that choice to the user. For example, connectivity assumptions tend to be embedded in each of the visual algorithms of `building block` libraries, a term used by Wongsuphasawat \cite{wongsuphasawatNavigatingWideWorld2021} to describe libraries that provide modular components for building elements of a visualization, such as functions for making boxes or translating data values to colors. In building block libraries such as Matplotlib\cite{hunterMatplotlib2DGraphics2007} and D3\cite{bostockDataDrivenDocuments2011} assumptions about connectivity are embedded in the interfaces such that the API is inconsistent across plot types. For example in Matplotlib methods for updating data and parameters for controlling aesthetics differ between (1D) line based plotting methods and (0D) marker based methods. While VTK\cite{hanwellVisualizationToolkitVTK2015,geveciVTK2012} provides a language for expressing the topological properties of the data, and therefore can embed that information in its visual algorithms, VTK's charts API is similar to the continuity dependent APIs of other building block libraries.

Domain specific libraries are designed with the assumption of continuities that are common in the domain \cite{HeerSoftware2006}, and therefore can somewhat restrict their API to choices that that are appropriate for the domain. For example, a tabular topological structure of discrete rows, as illustrated in \autoref{fig:related-work:continuity:ktypes}, is assumed by
A Presentation Tool\cite{mackinlayAutomatingDesignGraphical1986, mackinlayAutomatingDesignGraphical1986} and grammar of graphics\cite{wilkinsonGrammarGraphics2005} and the ggplot\cite{wickhamGgplot2ElegantGraphics2016}, vega\cite{satyanarayanDeclarativeInteractionDesign2014}, and altair\cite{vanderplasAltairInteractiveStatistical2018} libraries built on these frameworks. Image libraries such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and its humanities ImagePlot\cite{studiesCulturevisImageplot2021} plugin assume that the input is 2D continuous. Networking libraries such as gephi\cite{bastianGephiOpenSource2009} and networkx\cite{HagbergExploringNetwork2008} assume a graph-like structure. By assuming the structure of their data, these domain specific libraries can provide more cohesive interfaces for a much more limited set of visualization algorithms than the building block libraries offer.

@@

\section*{Acknowledgment}
\note{Acknowledge all the actual people}
The authors would like to thank the anonymous reviewers who gave constructive feedback on
an earlier version of this paper. Various Matplotlib and Napari contributors, particularly
Juan Nunez-Iglesias , and Nicolas Kruchten have provided valuable feedback from the
library developors perspective.

Hannah is also very grateful to Nicolas for the suggestion of augmented notation and to
the nlab and wikipedia contributors who wrote clear explanations of many of the topics
discussed in this paper.

This project has been made possible in part by grant number 2019-207333 and from the
\note{Cycle 3 grant #} Chan Zuckerberg Initiative DAF, an advised fund of Silicon Valley Community Foundation
